From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-28 Pi+/paper/ufin.tex
working on grpd fin

\section{The Universe of Finite Types}
\label{sec:ufin}

@@

The lambda calculus is classically presented with types, terms, and
$\beta\eta$-equations. This can be turned into a 1-category, with 0-cells for
types, and 1-cells for programs quotiented by congruence closure of the
$\beta\eta$-equations.

Then there is the free CCC, which is a 1-category constructed using
0-cells and 1-cells with universal properties that hold upto the notion
of equality that from the ambient metatheory.

Lambek/Scott show that these are equivalent. In modern language, this is
a 1-equivalence of (1,1)-categories.

Now, lets work without quotients in a proof-relevant metatheory.

The lambda calculus can now be presented with 0-cells for types, 1-cells
for terms, and 2-cells for equations/rewrites. This is a
(2,1)-category. Then, you can give a description of the free 2-CCC,
where universal properties hold upto canonical isomorphism.

Then we can show that there is a 2-equivalence of these
(2,1)-categories. (Actually, it can be weaker than (2,1)-categories, in
fact it holds for bicategories.)

Now, lets go to our result.

For Pi/Pi+, we present it as a (2,0)-category (or 1-groupoid à la HoTT),
with 0-cells for types, 1-cells for reversible programs, 2-cells for
rewrites/equations. And we present different descriptions (Pi+, Ufin)
of the free (2,0)-SMC (or 1-SMG) and show they're 2-equivalent.

So this is just a difference of presentation, in going from the (1,r) to
the (2,r) category. We have simply made the equations/rewrites
first-class in the presentation.

>>>>

\section{The (Sub)Universe of Finite Types}
\label{sec:ufin}

In the previous section, we have picked a particular encoding of finite types: as sequences of elements in some
canonical order.

This is not a ``free'' presentation. Some properties of the encoding might pollute our semantics. We want to make that
any theorems / properties of the semantics we prove apply to the source language (under various interpretations).

This motivates us to use a different semantic representation: a univalent universe of finite types which contains the
finite types and nothing but the finite types and is defined such as alternative representations of finite types are
identified. This guarantees that we can get a proper, sound and complete, semantics for our language.

\begin{center}
\begin{tikzpicture}[scale=1,every node/.style={scale=1}]
\usetikzlibrary{arrows}
\usetikzlibrary{shapes}
\node[ultra thick, draw=blue, ellipse, minimum width=300pt, minimum height=300pt, align=center] {};
\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (a) {$\{ 1 , 2 \} \uplus \{ 3 \}$};
\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (b) at ([shift=({70pt,-70pt})]a) {$\{ 1 \} \uplus \{ 2, 3 \}$};
\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (c) at
([shift=({-70pt,-70pt})]a) {$\{ 10 \} \uplus \{ 20, 30 \}$};
\node (d) at ([shift=({50pt,-10pt})]a) {$\cdots$};
\draw[thick,-] (a) to node[draw=none,right] {$\equiv$} (b);
\draw[thick,-] (b) to node[draw=none,below] {$\equiv$} (c);
\draw[thick,-] (a) to node[draw=none,left] {$\equiv$} (c);
\end{tikzpicture}
\end{center}

Given an operational model of how a type A reduces to a type B
Given a denotational model of which types A and B are equivalent
Want soundness and completeness results
Guarantee that the two notions of equality are themselves “equivalent”
Generalized “identity of indiscernibles”

@@
